---
title: '**BI 306 R Tutorial: Part 1**'
date: "*Fall 2021*"
output: 
  html_document:
    theme: cosmo
    highlight: tango
---

<br/>

Before you begin, you will need to download the following:   

* R software [here](https://archive.linux.duke.edu/cran/) for your computer   
* RStudio [here](https://www.rstudio.com/products/rstudio/download/#download)   

<br/>


Quick notes on this document:

* Blue text can be clicked on to access the associated link (for example, you can click [this](https://www.ipcc.ch) to link to the latest IPCC report)
* Text with grey shaded background represents R objects or commands in line with main text like this: ```2 + 2```


<br/>


## **Getting started with R**

To begin, we will be working with one of the built-in datasets in R called ```iris``` that contains the sepal and petal widths and lengths of different iris species ([you can read more about this dataset here](https://en.wikipedia.org/wiki/Iris_flower_data_set)). 

```{r}

## Load the iris dataset that is included in R
data("iris")

```

After loading the data, we can see what is actually included in the data using the ```head()``` function.

```{r}

head(iris)

```

The ```head()``` function will display the top six rows of data (you can modify this to show, for example, the top 3 rows like this ```head(iris, 3) ```). This is a nice way to get a sense of what your data look like. We can now see that the ```iris``` dataset contains four columns containing data for sepal length, sepal width, petal length, petal width, and species. These columns contain different types of data that we can understand a bit better by using the ```str()``` function to view the *structure* of the data.

```{r}

str(iris)

```

From this function, we can now see that the ```iris``` dataset is a ```data.frame``` object, it contains 150 observations (or rows of data), and 5 variables (columns). Additionally, the function displays the type of data in each column (identified after the ```$```) and first few observations. We see that the first four columns all contain numeric (```num```) values while the last column (```Species```) contains factors with three different levels (these levels are the different types of species within that column).

<br/>

## **Downloading and loading packages**

R comes with a lot of functions already included that you can use for your analyses, however, there are so many powerful and helpful packages out there that include additional functions that you will likely want to use. For example, we will want to calculate some simple summary statistics on the ```iris``` data and the package ```Rmisc``` contains a helpful function ```summarySE()``` that can accomplish this. In order to use this function, you will need to download this package. Most R packages can be easily downloaded in RStudio from the packages tab in the lower right panel or using the command ```install.packages()```. To practice, go ahead and download the ```Rmisc``` package using the command below.
```{r echo = TRUE, results = 'hide', eval = FALSE}

## Download the Rmisc package from CRAN
install.packages("Rmisc")

```

Once you have installed the package (*you only need to do this once or if you update your R version*), you need to load that package using the function ```library()``` or ```source()``` before you can use any of the included functions.
```{r message=FALSE}

## Load the Rmisc package before using it
library(Rmisc)

```

Great! You now have the ```Rmisc``` package installed and loaded in your current RStudio session. While you only need to install a package once, you will need to load the package every time you begin a new session and want to use those functions. 

<br/>

**Helpful tip**: If you are using a package for the first time or are looking for helpful information about the functions within a package, you can view the documentation for that package by running the following code with the name of the package you are interested in.
```{r}

## View the documentation associated with the Rmisc package
help(package = "Rmisc")

```

<br/>

## **Simply summary statistics**

Once you know what the structure ```str()``` of your data looks like, it is also good to look at some basic summary statistics. There are many different ways to calculate summary statistics in R, but here we will explore two different menothds using different R packages.

### **Method 1**: summarySE()
The first method will use the function ```summarySE()``` within the R package ```Rmisc``` installed and loaded above. 

### **Method 2**: Tidyverse
The second method we will use will rely on the ```tidyverse``` syntax. This is a really usefuly collection of R packages that are designed for data science and make your code very clean for other people to understand.  

Start by installing the ```tidyverse``` package.
```{r}

## You do this one! How do you install the tidyverse package? Add the code below:

```


<br/>

## **Helpful resources**

* [R Markdown cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [Data visualization with ggplot2 cheatsheet](https://hbctraining.github.io/Intro-to-R-flipped/cheatsheets/data-visualization-2.1.pdf)
* [Base R cheat sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)
* [A bunch of R cheat sheets](https://www.rstudio.com/resources/cheatsheets/)


