---
title: "R Assignment #3 Key"
date: '*Marine Biology: Spring 2024*'
output:
    word_document:  
      reference_docx: ../Misc_files/word_styles_ref.docx  
#    html_document:
#      theme: cosmo
#      highlight: tango
---

## **Third R Assignment:**

Using the data we have collected as a class for lab skills (```Lab_Skills_Data.xlsx```), you will complete the following tasks. For each figure you create, make sure you use descriptive X and Y axis labels with units where applicable and write an informative figure caption. All figures, captions, and tables should be saved and placed within a Word document to submit along with a brief description of your results.

1. Make a bar plot of average (+/- standard deviation) value per sample for each of the
following parameters:
    * pH
    * temperature
    * salinity
    * weights
    * calipers (L x W x H)
2. Make a scatter plot of the following parameters:
    * salinity versus temperature
    * shell width versus height
3. Find a new data visualization/figure to create in R to present some other relationship in your data that we have not learned yet in another R assignment (i.e., no line, scatter, or bar plots). Be creative and have fun!
4. Perform a statistical test in R (I recommend a T test or ANOVA) on one of the parameters you plotted in part 1 above. Present the results of your statistical test in a table.
5. Finally, write a brief results-style section discussing the statistical test you performed in part 4. In this section, present the p-value and discuss if your results were statistically different or not.

<br/>

---

<br/>

### Question 1: bar plot of average (+/- standard deviation)

```{r load libraries}

library(tidyverse)
library(readxl)

```

```{r load data}

# load temp/salinity data
temp_sal_df <- read_excel("~/Dropbox/Git/teaching_Rtutorial/Part_3/Lab_Skills_Data.xlsx", sheet = "Station 1 Salinity and Temp")

# load pH data
pH_df <- read_excel("~/Dropbox/Git/teaching_Rtutorial/Part_3/Lab_Skills_Data.xlsx", sheet = "Station 2 pH")

# load weight data
weight_df <- read_excel("~/Dropbox/Git/teaching_Rtutorial/Part_3/Lab_Skills_Data.xlsx", sheet = "Station 5 Byoyant Weight")

# load length/width data
calip_df <- read_excel("~/Dropbox/Git/teaching_Rtutorial/Part_3/Lab_Skills_Data.xlsx", sheet = "Station 4 Calipers")

```

<br/>


#### pH
```{r pH summary data}

# remove the first two rows of pH data (one is fake, the other is blank)
pH_df <- pH_df[-1:-2,]

# convert from wide to long to use all pH data
pH_long <- pH_df %>% 
  pivot_longer(names_to = "pH_meas", values_to = "pH", pH_1:pH_3) 

# summary of pH data by beaker ID
pH_sum <- pH_long %>% 
  group_by(Beacker_ID) %>% 
  summarise(pH_mean = mean(pH),
            pH_sd = sd(pH))

```

```{r pH bar plot}

ggplot(data = pH_sum, aes(x = Beacker_ID, y = pH_mean, fill = Beacker_ID)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = pH_mean - pH_sd, ymax = pH_mean + pH_sd)) +
  labs(x = "Beaker ID", y = "pH")

```

<br/>


#### temperature
```{r temp and sal long}

# remove the first two rows of temp data (one is fake, the other is blank)
temp_sal_df <- temp_sal_df[-1,] %>% 
  drop_na(Beacker_ID)

# convert from wide to long to use all temp data
temp_sal_long <- temp_sal_df %>% 
  pivot_longer(names_to = "temp_meas", values_to = "temp", Temp1:Temp3) %>% 
  pivot_longer(names_to = "sal_meas", values_to = "sal", Sal_probe_1:Sal_probe_3)

```

```{r temp summary data}

# summary of temp data by beaker ID
temp_sum <- temp_sal_long %>% 
  group_by(Beacker_ID) %>% 
  summarise(temp_mean = mean(temp),
            temp_sd = sd(temp))

```

```{r temp bar plot}

ggplot(data = temp_sum, aes(x = Beacker_ID, y = temp_mean, fill = Beacker_ID)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = temp_mean - temp_sd, ymax = temp_mean + temp_sd)) +
  labs(x = "Beaker ID", y = "Temperature (C)")

```

<br/>


#### salinity
```{r sal summary data}

# summary of sal data by beaker ID
sal_sum <- temp_sal_long %>% 
  group_by(Beacker_ID) %>% 
  summarise(sal_mean = mean(sal),
            sal_sd = sd(sal))

```

```{r sal bar plot}

ggplot(data = sal_sum, aes(x = Beacker_ID, y = sal_mean, fill = Beacker_ID)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = sal_mean - sal_sd, ymax = sal_mean + sal_sd)) +
  labs(x = "Beaker ID", y = "Salinity (PSU)")

```

<br/>


#### weights
```{r weight summary data}

# convert from wide to long to use all weight data
weight_long <- weight_df %>% 
  drop_na(weight_1) %>% 
  pivot_longer(names_to = "weight_meas", values_to = "weight", weight_1:weight_3)

# summary of weight data by beaker ID
weight_sum <- weight_long %>% 
  group_by(ID) %>% 
  summarise(weight_mean = mean(weight),
            weight_sd = sd(weight))

```

```{r weight bar plot}

ggplot(data = weight_sum, aes(x = ID, y = weight_mean, fill = ID)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = weight_mean - weight_sd, ymax = weight_mean + weight_sd)) +
  labs(x = "Item", y = "Weight (g)")

```

<br/>


#### calipers (L x W x H)
```{r calip summary data}

# convert from wide to long to use all calip data
calip_long <- calip_df[-1,] %>% 
  drop_na(Shell_ID) %>% 
  pivot_longer(names_to = "calip_meas", values_to = "measure", length_mm:height_mm)

# summary of calip data by beaker ID
calip_sum <- calip_long %>% 
  group_by(Shell_ID, calip_meas) %>% 
  summarise(calip_mean = mean(measure),
            calip_sd = sd(measure))

```

```{r calip bar plot}

ggplot(data = calip_sum, aes(x = Shell_ID, y = calip_mean, fill = Shell_ID)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = calip_mean - calip_sd, ymax = calip_mean + calip_sd)) +
  labs(x = "Shell ID", y = "Measure (mm)") +
  facet_wrap(~calip_meas)

```

<br/>


### Question 2:

```{r sal v temp}

ggplot(data = temp_sal_long, aes(x = sal, y = temp, color = Beacker_ID)) +
  geom_point() +
  labs(x = "Salinity (PSU)", y = "Temperature (C)")

```

```{r width v height}

ggplot(data = calip_df[-1,], aes(x = width_mm, y = height_mm, color = Shell_ID)) +
  geom_point() +
  labs(x = "Width (mm)", y = "Height (mm)")

```

<br/>


### Question 3:
Anything goes here! But this website has some great examples of different data visualization options: [https://r-graph-gallery.com/](https://r-graph-gallery.com/)

<br/>


### Question 4:

T test option:
```{r t test with table, message=FALSE}
library(knitr)

# comparing pH across two beakers
t.test(pH ~ Beacker_ID, data = pH_long)


# OR using the rempsyc package:
library(rempsyc)

ttest2 <- nice_t_test(data = pH_long, response = "pH", group = "Beacker_ID")
kable(ttest2)

```

<br/>


ANOVA option:
```{r anova with table}

summary(aov(pH ~ Group, data = pH_long))

```

<br/>


### Question 5:

Again, this section with vary but key details I am looking for here include:   

  * the parameter being tested (*e.g.*, pH, temperature, etc.)
  * statement of what statistical test was run 
  * the p value and interpretation of the p value
  * broad conclusion of what this may mean

<br/>


---

*This answer key was created by Colleen Bove, please feel free to update as needed for your own assignments!*
