---
title: 'Marine Biology R Assignment #4'
date: '*Spring 2024*'
output:
  word_document: default
  html_document:
    theme: cosmo
    highlight: tango
---

```{r eval=FALSE, include=FALSE}

# Need to install the following (for Mac) if knitting to PDF
tinytex::install_tinytex()

```

## Fourth R Assignment

**This R assignment will be due by 3PM on Tuesday, March 26th**. You will submit both your R markdown code and the Word output of your work. This assignment can be completed with other members from your lab section but if you work with others, please include their names on your submission. 

Using the ```MarineBio_ExampleData.csv``` data that can be found on Canvas, you will complete the following tasks:

1. **Create a R markdown document that will knit as a Word document.** You will submit both the R markdown code and the Word document for this assignment so this will be assessed as a part of your submission. 
2. **Recreate <span style="color:teal">Figure 1</span> (below) as closely as possible** See if you can figure out what data are shown in this figure and then ask for help if you are unsure. Using what you have learned about R so far, try to recreate the figure with captions, labels, colors, shapes, etc. as closely as possible. 
3. **Recreate <span style="color:teal">Figure 2</span> (below) as closely as possible** See if you can figure out what data are shown in this figure and then ask for help if you are unsure. Using what you have learned about R so far, try to recreate the figure with captions, labels, colors, line types, etc. as closely as possible. 
4. **Perform an Analysis of Variance (ANOVA) of the effect of *p*CO~2~ on calcification rate (```rate```)** Show your code for how you performed your statistical test and then have your markdown file report your p-values from your different treatment. Include a brief (~2-3 sentences) interpreting your results from your ANOVA citing statistics in your statement. 

```{r}

df <- read.csv("./Part_4/MarineBio_ExampleData.csv", header = TRUE)
str(df)

```

```{r}

library(tidyverse)

```

---

## Question 1:

For full credit, you should submit both the Rmarkdown (.Rmd) and resulting Word (.doc or .docx) documents demonstrating your code and the resulting output. 

<br/>


## Question 2: Figure 1

```{r lipid summary for plot}

lipid_sum <- df %>% 
  group_by(treat, species) %>% 
  summarise(mean = mean(lipid),
            sd = sd(lipid))

```

```{r lipid figure}

ggplot(data = lipid_sum, aes(x = treat, colour = species)) +
  geom_point(aes(y = mean), shape = 8, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), position = position_dodge(width = 0.5)) +
  scale_colour_brewer("Coral Species", labels = c("PAST", "PSTR", "SSID", "UTEN"), palette = "Dark2") +
  labs(x = "Treatment",
       y = "Lipids (mg/cm2)",
       title = "Coral Lipid Content Under Ocean Acidification and Warmging",
       subtitle = "Mean +/- standard deviation") +
  theme_classic()

```

<br/>


## Question 3: Figure 2

```{r remove T0 samples}

T90_df <- df %>% 
  filter(T0_T90 == "T90")

```

```{r rate v den figure}

ggplot(data = T90_df, aes(x = rate, y = den, colour = ftemp, linetype = ftemp)) +
  geom_point(alpha = 0.5) +
  scale_color_manual("Temperature", values = c("blue", "red"), labels = c("28 C", "31 C")) +
  scale_linetype_manual("Temperature", values = c("dotted", "dashed"), labels = c("28 C", "31 C")) +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~ species, scales = "free_y", labeller = as_labeller(c("A" = "PAST", "P" = "PSTR", "S" = "SSID", "T" = "UTEN"))) +
  labs(x = "Calcification Rate (mg/cm2/day)",
       y = "Symbiont Density (cells/cm2)",
       title = "Coral calcification rate versus chlorophyll a content") +
  theme_bw()

```

<br/>


## Question 4:

```{r rate v pco2 anova}

# run a 1-way ANOVA of the effect of pCO2 treatment on calcification rate
anova_out <- aov(rate ~ pco2, data = df)

# view summary of ANOVA model output
summary(anova_out)

```

**Results:** After performing a one-way analysis of variance to assess the effects of *p*CO~2~ treatment on coral calcification rate, we can determine that *p*CO~2~ treatment significantly alters coral calcification rate (F = 41.55; p < 0.001). We thus reject the null hypothesis in favor of the alternate hypothesis that our experimental treatment had a significant effect on our response variable.


<br/>


---

*This answer key was created by Colleen Bove, please feel free to update as needed for your own assignments!*
